---
title: "Metadata audit for reviewer-requested bias checks"
author: "Chao-Chung Kuo"
format:
  html:
    toc: true
    code-fold: true
    embed-resources: true
    theme: cosmo
    css: metadata_report.css
    page-layout: full
    smooth-scroll: true
execute:
  echo: true
  warning: false
  message: false
---

This report validates the mapping between Juliane's anonymized reviewer-round
metadata and the 73 samples in the post-QC methylation matrix. The supplied Excel
table is treated as authoritative for this analysis.

The `lymphomonocytes` field is reproduced exactly as supplied. Its scoring method
and biological meaning are unknown to the analyst; it is therefore not interpreted
and is not used as an adjustment covariate. Requested variables absent from the
supplied table are not inferred.

```{r}
source("../../R/config.R")
source("../../R/metadata.R")
cfg <- read_project_config("../..")
load(file.path(cfg$paths$project_data_root,
  "analysis/DNAmArray/DNAmArray_Processing.RData"))
bias <- read_bias_metadata(cfg$paths$bias_metadata)
key <- read_sample_key(cfg$paths$sample_sheet)
metadata <- build_analysis_metadata(bias, key, colnames(mVals))
validation <- validate_analysis_metadata(metadata, colnames(mVals))
validation
```

## Metadata overview

```{r}
knitr::kable(as.data.frame(table(metadata$display_group)),
  col.names = c("Disease group", "N"))
```

## Cross-tabulations

```{r, results='asis'}
tabs <- make_cross_tabs(metadata)
for (nm in names(tabs)) {
  cat("\n### Disease group by ", nm, "\n\n", sep = "")
  if (nm == "sentrix_id") {
    # Put the many Sentrix IDs in rows rather than columns. This preserves the
    # complete cross-tabulation while keeping the table within the report width.
    sentrix_counts <- as.matrix(tabs[[nm]][, -1, drop = FALSE])
    storage.mode(sentrix_counts) <- "numeric"
    rownames(sentrix_counts) <- tabs[[nm]]$disease_group
    sentrix_matrix <- t(sentrix_counts)
    sentrix_table <- data.frame(
      `Sentrix ID` = rownames(sentrix_matrix),
      sentrix_matrix,
      check.names = FALSE,
      row.names = NULL
    )
    sentrix_table$Total <- rowSums(sentrix_matrix)
    cat('<div class="table-scroll sentrix-table-wrap">\n')
    print(knitr::kable(
      sentrix_table,
      format = "html",
      row.names = FALSE,
      align = c("l", rep("c", ncol(sentrix_table) - 1)),
      table.attr = 'class="table sentrix-matrix"'
    ))
    cat('\n</div>\n')
  } else {
    cat('<div class="table-scroll">\n')
    print(knitr::kable(tabs[[nm]]))
    cat('\n</div>\n')
  }
  cat("\n")
}
```

## Interpretation boundary

These tables describe structural overlap between disease groups and supplied
metadata. They do not by themselves establish causality or demonstrate that a
variable is a biological confounder. Where disease group and a metadata category
are perfectly or nearly perfectly aligned, their effects may not be statistically
separable in this cohort.
